The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-07-07 paper/ms.tex
new edits in discussion section

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions in the sphere: central and a
homogeneous.
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle and rotational velocity. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis both the intensity at the line
center and the line width increase with rotational velocity.   
Along the same line of sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux, the escape fraction or the
average number of scatterings. From this, we conclude that the \lya scattering
through a rotating solid-body proceeds identical to the static
case. The only difference is the doppler shift from the different
regions in the sphere that move with respect to the observer. This
allows us to derive an analytic approximation for the viewing-angle
and rotational velocity of the emerging spectrum.
\end{abstract}

@@

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results. In
the last section we present our conclusions. The Appendix presents the
derivation of an analytic expression that can be helpful to interpret
to first order the main trends observed in the Monte Carlo simulations.

@@

Here we briefly describe the relevant characteristics of the two
radiative transfer codes we have used.  
For a detailed description we refer the reader to the original papers
\cite{CLARA,DijkstraKramer}.

@@

The initialization process for the \ly photons specifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the gas reference frame and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

@@

The codes treat the gas as homogeneous in density and temperature.  
This implies that the cas is completely defined by its geometry
(i.e. sphere or slab), temperature $T$, Hydrogen optical depth
$\tau_{\rm H}$, dust optical depth $\tau_{\rm a}$ and the bulk
velocity field $\vec{v}$.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. 
For each galaxy we vary at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. 
In total there are $48$ different models combining all the possible
different variations in the input parameters.
Table \ref{table:models} lists the different parameters we used to
generate the models.

Additionally, we have used two independently developed Monte Carlo
codes \citep{CLARA,DijkstraKramer} to perform the calculations. 
The results and trends we report are observed in both codes.

@@

The central result of this paper are summarized in Fig.s
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parametrized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have verified that the solutions are indeed symmetric with respect
to $\mu=0$.

From these Fig.s we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximum, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity.

@@

The first column in both Fig.s \ref{fig:CentralSpec} and
\ref{fig:HomSpec} shows that for the static sphere the line properties
are independent of $|\mu|$, as it is expected due
to the spherical symmetry. 
However, for increasing rotational velocities, at a fixed optical
depth, there are clear signs that this symmetry is broken.

@@

This is clear in  Fig. \ref{fig:differentvelocities} where we
present the different line morphologies for $|\mu|\sim 0$ in the
homogeneous and central configurations.  
The panels have the same distribution as Fig.s \ref{fig:CentralSpec}
and \ref{fig:HomSpec}.    
There are three clear effects on the line morphology as the rotational
velocity increases.  
First, the line broadens; second, the double peaks reduce their intensity; and
third, the intensity at the line centre rises. 
The last two effects are combined to give the impression that the double
peaks are merged into a single one at high rotational velocities.

@@

\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation} 
%
where $\mu=\cos\theta$, $N$ is the total number of outgoing photons,
$\Delta N$ is the number of photons in an angular bin $\Delta
\Theta$. This definition satisfies the condition
$\int_{-1}^{1}F(\mu)d\mu/2=1$.  In the case of perfect spherical
symmetry one expects a flat distribution with $F(\mu)=1$.

Fig. \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and the two
types of source distributions.
This Fig. shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of \%2.

This is a remarkable result. 
In spite of that the rotation axis defines preferential direction, the
integrated flux is the same for all viewing angles, at least for the
range of parameters explored in this paper.

@@

In the optically thick regime, \lya  photons can escape in
'single flight' which corresponds to a scenario in which the
photon resonantly scatters $10^4-10^5$ times until it is scattered
into the wing of the line ($x\sim 3-4$). 
At these frequencies the medium is optically thin, and the photons can
escape efficiently in a single flight. 
In contrast, in an extremely optically thick medium \ly
  photons escape in a `single excursion' \citep{Adams72}. 
Here, photons that are scattered into the wing of the line escape from
the medium in a sequence of wing scattering events. 
In both cases, \ly photons resonantly scatter $10^4$-$10^5$ times. 
Because we keep our clouds the same size, the mean free path of Lya
photons that scatter resonantly is 10 times larger for the case
$\tau_{\rm H}=10^5$ than for $\tau_{\rm H}=10^6$. 
If we compute the average distance $D$ travelled by \ly
photons through a medium of size $R$ as a function of line center
optical depth $\tau_{\rm H}$, then we find that during the transition
from optically thick to extremely optically thick the mean traversed
distance $D$ actually decreases slightly.  
This decrease is unique to this transition region, and $D$ generally
increases with $\tau_{\rm H}$ at other values of $\tau_{\rm H}$.

@@

One way to understand this is that a rotating sphere
is practically identical from a static one. 
In the frame of the rotating gas cloud all atoms are in fact
stationary with respect to each other, making the scattering process
proceed identical as in the static case.  
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each other and the scattering process
proceeds identical as in the static case.

@@

{\bf Galaxy simulations with gas rotation}. \cite{Verhamme12} studied \ly
line emission in two high resolution simulations of individual
galaxies. 
The main purpose of their study was to assess the impact of two
different Inter-Stellar Medium (ISM) prescriptions. 
However, each simulated galaxy had a disc structure with a clear rotation pattern in
the ISM and inflowing gas from the circum-galactic region. 
The configuration had an axial symmetry and they reported a strong dependence of both
the escape fraction and the total line intensity as a function of the
$\theta$ angle. 
From our study, none of these two quantities has a dependence either
on the inclination angle or the rotational velocity. 
We suggest that he effect reported by \cite{Verhamme12} is
consistent with being a consequence of the different hydrogen optical
depth for different viewing angles and not as an effect of the bulk
rotation.

@@

{\bf Single peaked lines}. The presence of single peaked profiles has
been associated to inflow/outflow dynamics
\citep{Verhamme06,DijkstraKramer}. 
Gas bulk rotation can also be considered as a probable origin for that
behaviour, provided that the observed single peak is highly
symmetric. 
Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be
considered in the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. 
This highlights
that in order to interpret the observations of \ly emitting galaxies
it is necessary to consider the possible effect of different
inclination angles along the line of sight.

{\bf Systemic velocities}. There are observational measurements for the
velocity shift between the \ly and other emission lines. In our study
we find that the position of the peak maxima is insensitive to the
rotational velocity of the gas. In our model the only parameter that
can influence the peak maxima position is the optical depth.

@@

\item In all of our models, rotation induces changes in the line morphology
 for different values of the angle between the rotation
 axis and the LoS, $\theta$.  The changes are such that for 
 $\cos\theta\equiv \mu\sim 0$, i.e. LoS perpendicular to the
 rotation axis, and high rotational velocities the line becomes single peaked.

\item For a viewing angle with $\mu=0$ the line width increases with rotational
  velocity. This change approximately follows the functional form  ${\rm FWHM}^2 = {\rm FWHM}_{ 0}^2 + (V_{\rm max}/\lambda)^2$, where FWHM$_{0}$ indicates the line
  width for the static case and $\lambda$ is a constant. We have
  determined this constant to be  $\lambda_{\rm c}=0.83 \pm 0.06$ and
  $\lambda_{\rm h}=0.82\pm 0.05$ for the central and homogeneous source
  distributions, respectively.

@@

The definitive and clear impact of rotation on the \ly morphology
suggests that this is an effect that should be taken into account at
the moment of interpreting high resolution spectroscopic data. To this
end, we defer a discussion of observed lines features in the context of gas
bulk rotation to a future publication (Remolina-Gutierrez et al., in
prep.).

>>>>

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions in the sphere: central and 
homogeneous.
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle and rotational velocity. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis both the intensity at the line
center and the line width increase with rotational velocity.   
Along the same line of sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux, the escape fraction or the
average number of scatterings. From this, we conclude that the \lya scattering
through a rotating solid-body proceeds identical to the static
case. The only difference is the doppler shift from the different
regions in the sphere that move with respect to the observer. This
allows us to derive an analytic approximation for the viewing-angle
and rotational velocity of the emerging spectrum.
\end{abstract}

@@

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results. In
the last section we present our conclusions. The Appendix presents the
derivation of an analytic expression to interpret
the main trends observed in the Monte Carlo simulations.

@@

Here we briefly describe the relevant characteristics of the two
radiative transfer codes we have used.  
For a detailed description we refer the reader to the original papers
\cite{CLARA} and \cite{DijkstraKramer}.

@@

The initialization process for the \ly photons specifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the gas reference frame and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{\rm in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

@@

The codes treat the gas as homogeneous in density and temperature.  
This implies that the gas is completely defined by its geometry
(i.e. sphere or slab), temperature $T$, Hydrogen optical depth
$\tau_{\rm H}$, dust optical depth $\tau_{\rm a}$ and the bulk
velocity field $\vec{v}$.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. 
For each galaxy we vary at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. 
In total there are $48$ different models combining all the possible
different variations in the input parameters.
Table \ref{table:models} lists the different parameters we used to
generate the models. The results and trends we report are observed in both 
Monte Carlo codes.

@@

The main results of this paper are summarized in Fig.
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parametrized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have verified that the solutions are indeed symmetric with respect
to $\mu=0$.

From these figures we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximum, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity and viewing angle.

@@

The first column in both Fig. \ref{fig:CentralSpec} and
\ref{fig:HomSpec} shows that for the static sphere the line properties
are independent of $|\mu|$, as it is expected due
to the spherical symmetry. 
However, for increasing rotational velocities, at a fixed optical
depth, there are clear signs that this symmetry is broken.

@@

This is clear in  Fig. \ref{fig:differentvelocities} where we
present the different line morphologies for $|\mu|\sim 0$ in the
homogeneous and central configurations.  
The panels have the same distribution as Fig. \ref{fig:CentralSpec}
and \ref{fig:HomSpec}.    
There are three clear effects on the line morphology as the rotational
velocity increases.  
First, the line broadens; second, the double peaks reduce their intensity; and
third, the intensity at the line centre rises. 
The last two effects are combined to give the impression that the double
peaks are merged into a single one at high rotational velocities.

@@

\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation} 
%
where $\mu=\cos\theta$, $N$ is the total number of outgoing photons,
$\Delta N$ is the number of photons in an angular bin $\Delta
\theta$. This definition satisfies the condition
$\int_{-1}^{1}F(\mu)d\mu/2=1$.  In the case of perfect spherical
symmetry one expects a flat distribution with $F(\mu)=1$.

Fig. \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and the two
types of source distributions.
This shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of \%2.

This is a remarkable result. 
In spite of that the rotation axis defines preferential direction, the
integrated flux is the same for all viewing angles in the
range of parameters explored in this paper.

@@

In the optically thick regime, \ly  photons can escape in
'single flight' which corresponds to a scenario in which the
photon resonantly scatters $10^4-10^5$ times until it is scattered
into the wing of the line ($x\sim 3-4$). 
At these frequencies the medium is optically thin, and the photons can
escape efficiently in a single flight. 
In contrast, in an extremely optically thick medium \ly
  photons escape in a `single excursion' \citep{Adams72}. 
Here, photons that are scattered into the wing of the line escape from
the medium in a sequence of wing scattering events. 
In both cases, \ly photons resonantly scatter $10^4$-$10^5$ times. 
Because we keep our clouds the same size, the mean free path of Lya
photons that scatter resonantly is 10 times larger for the case
$\tau_{\rm H}=10^5$ than for $\tau_{\rm H}=10^6$. 
If we compute the average distance $D$ travelled by \ly
photons through a medium of size $R$ as a function of line center
optical depth $\tau_{\rm H}$, then we find that during the transition
from optically thick to extremely optically thick the mean traversed
distance $D$ actually decreases slightly.  
This decrease is unique to this transition region, and $D$ generally
increases with $\tau_{\rm H}$ at other values of $\tau_{\rm H}$.

@@

One way to understand this is that a rotating sphere
is identical from a static one. 
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each other and the scattering process
proceeds identical as in the static case.

@@

{\bf Galaxy simulations with gas rotation}. \cite{Verhamme12} studied \ly
line emission in two high resolution simulations of individual
galaxies. 
The main purpose of their study was to assess the impact of two
different ISM prescriptions. 
However, each simulated galaxy had a disc structure with a clear rotation pattern in
the ISM and inflowing gas from the circum-galactic region. 
The configuration had an axial symmetry and they reported a strong dependence of both
the escape fraction and the total line intensity as a function of the
$\theta$ angle. 
From our study, none of these two quantities has a dependence either
on the inclination angle or the rotational velocity. 
We suggest that he effect reported by \cite{Verhamme12} is
consistent with being a consequence of the different hydrogen optical
depth for different viewing angles and not as an effect of the bulk
rotation.

@@

{\bf Single peaked lines}. The presence of single peaked profiles has
been associated to inflow/outflow dynamics
\citep{Verhamme06,DijkstraKramer}. 
Gas bulk rotation can also be considered as a probable origin for that
behaviour, provided that the observed single peak is highly
symmetric. 
Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be
considered in the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. 
This highlights
that in order to interpret the observations of \ly emitting galaxies
it is necessary to consider the possible effect of rotation and different
inclination angles along the line of sight.

{\bf Systemic velocities}. There are observational measurements for the
velocity shift between the \ly and other emission lines. In our study
we find that the position of the peak maxima can suddenly change with
rotation and viewing angle. Namely the line can become single peaked
for high rotational velocities and viewing angles perpendicular to the
rotation axis. This highlights the importance of considering rotation
in the interpretation of systemic velocity observations.

@@

\item In all of our models, rotation induces changes in the line morphology
 for different values of the angle between the rotation
 axis and the LoS, $\theta$.  The changes are such that for 
 a viewing angle perpendicular to the
 rotation axis, and high rotational velocities the line becomes single peaked.

\item The line width increases with rotational
  velocity. For a viewing angle perpendicular to the rotation axis
  This change approximately follows the functional form  ${\rm FWHM}^2
  = {\rm FWHM}_{ 0}^2 + (V_{\rm max}/\lambda)^2$, where FWHM$_{0}$
  indicates the line 
  width for the static case and $\lambda$ is a constant. We have
  determined this constant to be  $\lambda_{\rm c}=0.83 \pm 0.06$ and
  $\lambda_{\rm h}=0.82\pm 0.05$ for the central and homogeneous source
  distributions, respectively.

@@

The definitive and clear impact of rotation on the \ly morphology
suggests that this is an effect that should be taken into account at
the moment of interpreting high resolution spectroscopic data. In
particular it is relevant to consider the joint effect of rotation the
and ubiquitous outflows (Remolina-Gutierrez et al., in prep.).
